---
title: "STAT/MATH 495: Problem Set 09"
author: "Jeff Lancaster"
date: "2017-11-07"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 2
    collapsed: false
    smooth_scroll: false
    df_print: kable
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, fig.width=8, fig.height=4.5, message=FALSE, warning = FALSE
  )
set.seed(76)


library(stats)
library(mosaic)
library(ggplot2)
library(readr)
```



# Collaboration

Please indicate who you collaborated with on this assignment: 



# Question 1: Run k-means

```{r}
observations_1 <- read_csv("data/observations_1.csv")
observations_2 <- read_csv("data/observations_2.csv")

# Set observations to be one of two datasets
observations <- observations_2

# Fit model for k=2
k <- 2
k_means_results <- kmeans(observations, centers=k)
clusters <- k_means_results$cluster
cluster_centers <- k_means_results$centers

# Add cluster results to observations. Note we convert to factor since cluster
# ID's should be treated as categorical
observations$cluster <- as.factor(clusters)

# Add cluster ID's to cluster_centers
cluster_centers <- cluster_centers %>% 
  as_tibble() %>% 
  mutate(cluster=as.factor(1:k))

ggplot(NULL, aes(x=x1, y=x2, col=cluster)) +
  geom_point(data=observations) +
  geom_point(data=cluster_centers, size=5)
```

**Questions**:

1. Run KMC 10 times on `observations_1` and comment on the consistency of the
results.

```{r}
#simply runs k-means on a dataset
cluster <- function (k, data){
  output <- kmeans(data, centers=k)
  clusters <- output$cluster
  centers <- output$centers
}

clusters<-do(10)*cluster(2,observations_1)

ggplot(clusters,aes(x=x1,y=x2))+geom_point()
```
 
The results don't seem to be consistent at all.  As you can see in the points above, the less variance there is between sets of points, the more consistency there is - and there is no central location that the points are centered around.


2. Speculate on the root cause of any consistency or inconsistency in the
results.

```{r}
ggplot(observations_1,aes(x=x1,y=x2))+geom_point()
```

Looking at the outlier points in particular (the ones that are furthest away from the other 3 of their cluster), I am willing to guess that they are the result of a couple of points (or possibly even one) being sorted into the wrong cluster.  We can further this claim by looking at the plot of the observations_1 dataset and observing that there are no meaningful clusters present.

3. Run KMC 10 times on `observations_2` and comment on the consistentcy of the
results.

```{r}
clusters2<-do(100)*cluster(2,observations_2)

ggplot(clusters2,aes(x=x1,y=x2))+geom_point()

```

Whoa! The cluster centers are completely consistent - they are exactly the same in each instance.  I didn't believe it, so I increased the n to 100, and got the same result: perfectly consistent.

4. Speculate on the root cause of any consistency or inconsistency in the
results.

```{r}
ggplot(observations_2,aes(x=x1,y=x2))+geom_point()
```

The data is already broken up into 2 clusters!  This eliminates randomness when we call kmeans() on our dataset and causes the centers to be the same every time.


# Bonus question: Code your own

Read ISLR page 388 Algorithm 10.1 and implement k-means clustering from scratch.
Don't worry about doing it for general $k$; keep it simple and do it for $k=2$
specifically. Apply it to `observations_2` from above.

```{r}
# Hint:
library(proxy)
A <- data_frame(
  x1 = c(0, 0.5, 0.75, 1),
  x2 = c(0, 0.5, 0.75, 1)
)
B <- data_frame(
  x1 = c(1, 0),
  x2 = c(1, 0)
)
distance_matrix <- proxy::dist(x=A, y=B)
distance_matrix
apply(distance_matrix, 1, which.min)
```
